---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Natalie Holsclaw"
date: "Spring 2025"
output:
  pdf_document: default
  html_document:
    df_print: paged
geometry: margin=2.54cm
editor_options:
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
# Load necessary packages
library(tidyverse)
library(agricolae)
library(here)

# Check working directory
getwd()

# Import data
ntl_data <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"),
                     stringsAsFactors = TRUE)

# Convert dates to date format
ntl_data$sampledate <- as.Date(ntl_data$sampledate, format = "%m/%d/%y")

#2 Build ggplot theme
mytheme <- theme(panel.background = element_rect(fill = "cornsilk"),
        legend.title = element_text(color = "darkseagreen4"),
        axis.text.x = element_text(angle = 45),
        panel.grid.major = element_line(color = "white"))
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: The mean lake temperatures don't vary across depths in July.
Ha: The mean lake temperature do vary across depths in July. 


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
# Adding a month column
ntl_data_v2 <- mutate(ntl_data, month = month(sampledate)) 

# Filter data
ntl_data_v3 <- ntl_data_v2 %>% 
  filter(month == 7) %>% 
  select(lakename:daynum, depth:temperature_C) %>% 
  na.omit()

#5
# Visualize the data
plot1 <- ggplot(ntl_data_v3, aes(x = depth, y = temperature_C))+
  geom_point()+
  geom_smooth(method = "lm")+
  ylim(0,35)+
  labs(x = "Depth (m)", y = "Temperature (\u00B0C)",
       title = "Temperature vs Lake Depths in July")

# View plot
plot1
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: It does appear that temperature decreases with increasing depth meaning that there may be a negative linear relationship.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
regression <- lm(data = ntl_data_v3, temperature_C ~ depth)

summary(regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: There is a significant relationship between temperature and depth (df = 9726, p < 0.001). About 74% of variability in temperature is explained by changes in depth (R^2 = 0.7387). For every 1 m change in depth, we expect temperature to decrease by 1.94 degrees celsius. 


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

aic_test <- lm(data = ntl_data_v3, temperature_C ~ year4 + daynum + depth)
step(aic_test)

#10

summary(aic_test)
```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method did not exclude any of the variables (year4, daynum, depth). This model explains 74% of the observed variance (R^2 = 0.7411). This is a slight improvement from the model using depth as the only explanatory variable. 



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

# ANOVA with aov() function
anova1 <- aov(data = ntl_data_v3, temperature_C ~ lakename)
summary(anova1)

# ANOVA with lm() function
anova2 <- lm(data = ntl_data_v3, temperature_C ~ lakename)
summary(anova2)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, there is a significant difference in mean temperature among the lakes (p < 0.001).



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

fig1 <- ggplot(ntl_data_v3, aes(x = depth, y = temperature_C, color = lakename))+
  geom_point(alpha = 0.50)+
  geom_smooth(method = "lm", se = FALSE)+
  ylim(0,35)+
  labs(x = "Depth (m)", y = "Temperature (\u00B0C)",
       title = "Temperature vs Lake Depths in July")

fig1

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15

TukeyHSD(anova1)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Paul Lake and Ward Lake are statistically the same as Peter Lake as far as mean temperatures in July go. The mean temperature at Peter Lake is significantly different from Central Long Lake, Crampton Lake, East Long Lake, Hummingbird Lake, Tuesday Lake, and West Long Lake. There is no lake the is statistically distinct from all the other lakes.

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We could perform a t-test to see whether they have distinct mean temperatures. 



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
ntl_data_filterd <- ntl_data_v3 %>% 
  filter(lakename == "Crampton Lake" | lakename == "Ward Lake")

t_test_ntl <- t.test(ntl_data_filterd$temperature_C ~ ntl_data_filterd$lakename)
t_test_ntl # view results
```

>Answer: The mean temperatures in Crampton and Ward Lakes are not significantly different (p = 0.2649). The mean temperature for Crampton Lake is 15.35 C while the mean temperature for Ward Lake is 14.46 C. This does match the results of the Tukey HSD test. 
